Combined annotation exploration for SCPCP000006

### 2024-10-31 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
update code to create a final table of annotations

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE,
                      warnings=FALSE,
                      knitr.digits.signif = FALSE)

```

@@

@@

```{r path_to_query}
sample_metadata_file <- file.path(repository_base, "data", "current","SCPCP000006", "single_cell_metadata.tsv")

@@

# Create a data frames of all annotations
cell_type_df <- metadata$scpca_sample_id |>
  purrr::map(
    # For each sample_id, do the following:
    \(sample_id) {

      # Read in the Seurat object
     if(file.exists(file.path(result_dir, sample_id, paste0("06_infercnv_HMM-i3_", sample_id, "_reference-both.rds")))){

       srat <- readRDS(
        file.path(result_dir, sample_id, paste0("06_infercnv_HMM-i3_", sample_id, "_reference-both.rds"))
      )

@@

        # cnv global estimation per chromosome
        has_cnv_chr1 = srat$has_cnv_chr1,
        has_cnv_chr2 = srat$has_cnv_chr2,
        has_cnv_chr3 = srat$has_cnv_chr3,
        has_cnv_chr4 = srat$has_cnv_chr4,
        has_cnv_chr5 = srat$has_cnv_chr5,
        has_cnv_chr6 = srat$has_cnv_chr6,
        has_cnv_chr7 = srat$has_cnv_chr7,
        has_cnv_chr8 = srat$has_cnv_chr8,
        has_cnv_chr9 = srat$has_cnv_chr9,
        has_cnv_chr10 = srat$has_cnv_chr10,
        has_cnv_chr11 = srat$has_cnv_chr11,
        has_cnv_chr12 = srat$has_cnv_chr12,
        has_cnv_chr13 = srat$has_cnv_chr13,
        has_cnv_chr14 = srat$has_cnv_chr14,
        has_cnv_chr15 = srat$has_cnv_chr15,
        has_cnv_chr16 = srat$has_cnv_chr16,
        has_cnv_chr17 = srat$has_cnv_chr17,
        has_cnv_chr18 = srat$has_cnv_chr18,
        has_cnv_chr19 = srat$has_cnv_chr19,
        has_cnv_chr20 = srat$has_cnv_chr20,
        has_cnv_chr21 = srat$has_cnv_chr21,
        has_cnv_chr22 = srat$has_cnv_chr22
        ) |>
        tibble::rownames_to_column("barcode") |>
        dplyr::mutate(sample_id = sample_id)
     }
    }
  ) |>
  # now combine all dataframes to make one big one
  dplyr::bind_rows()

```

@@

```{r fig.width=10, fig.height=4, out.width='100%'}
do_Feature_mean <- function(df, group.by, feature ){

  df <- df %>%
  group_by(sample_id, !!sym(group.by))%>% 
  summarise(m = mean(!!sym(feature)))

p <- ggplot(df, aes(x= sample_id, y = !!sym(group.by), fill = m)) +
   geom_tile() +
   scale_fill_viridis_c()  +
   theme_bw() +
   theme(text = element_text(size = 20)) +
   theme(axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 0.5)) +
   guides(fill=guide_colourbar(title=paste0(feature)))

@@

```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}

cell_type_df <- cell_type_df |>
  mutate(has_cnv_score = rowSums(cell_type_df[,grepl("has_cnv_chr", colnames(cell_type_df))])) |>
  mutate(has_cnv_score = case_when(
    has_cnv_score > params$cnv_threshold ~ "CNV",
    has_cnv_score <= params$cnv_threshold ~ "no CNV"
  ))

@@

```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}

# Define normal cells
# We first pick up the immune and endothelial cells annotated via the label transfer compartments under the condition that the predicted score is above the threshold
cell_type_df <- cell_type_df |>
  mutate(first.level_annotation = case_when(
    # assign normal/cancer based on condition
    compartment %in% c("fetal_nephron", "stroma") & has_cnv_score == "no CNV" ~ "normal",
    compartment %in% c("endothelium", "immune") & cell_type.score > params$predicted.celltype.threshold ~ "normal",
    compartment %in% c("fetal_nephron", "stroma") & has_cnv_score == "CNV" ~ "cancer",
    .default = "unknown"

  ))

```

@@

```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
ggplot(cell_type_df, aes( x = umap.umap_1, y = umap.umap_2, color = first.level_annotation), shape = 19, size = 1)+
  geom_point() +
  facet_wrap(facets = ~sample_id, ncol = 5) +
  theme_bw()  +
  theme(text = element_text(size=22))

```

@@

   # assign cancer cells based on condition
   cell_type_df$compartment %in% c("stroma") &
          cell_type_df$has_cnv_score == "CNV"                         ~ "cancer stroma",

          cell_type_df$compartment %in% c("fetal_nephron") &
          cell_type_df$has_cnv_score == "CNV" &
          cell_type_df$cell_type == "mesenchymal cell"                ~ "blastema",

          cell_type_df$compartment %in% c("fetal_nephron") &
          cell_type_df$has_cnv_score == "CNV" &
          cell_type_df$cell_type != "mesenchymal cell"                ~ "cancer epithelium",
    .default = "unknown"

  ))

```

```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
ggplot(cell_type_df[cell_type_df$first.level_annotation == "normal",], aes( x = umap.umap_1, y = umap.umap_2, color = second.level_annotation), shape = 19, size = 1)+
  geom_point() +
  facet_wrap(facets = ~sample_id, ncol = 5) +
  theme_bw()  +
  theme(text = element_text(size=22))
```

```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
ggplot(cell_type_df[cell_type_df$first.level_annotation == "cancer",], aes( x = umap.umap_1, y = umap.umap_2, color = second.level_annotation), shape = 19, size = 0.1)+
  geom_point() +
  facet_wrap(facets = ~sample_id, ncol = 5) +
  theme_bw() +
  theme(text = element_text(size=22))
```

@@

```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
ggplot(cell_type_df, aes( x = umap.umap_1, y = umap.umap_2, color = second.level_annotation), shape = 19, size = 1)+
  geom_point() +
  facet_wrap(facets = ~sample_id, ncol = 5) +
  theme_bw()  +
  theme(text = element_text(size=22))
```

@@

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
  for(i in 1:22){
    print(do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = glue::glue("proportion_cnv_chr", i)))
  }

```

@@

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}

    do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000081237")

```

@@

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}

    do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000110799")

```

@@

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}

    do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000026025")

```

@@

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}

    do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000163359")

```

@@

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}

    do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000154096")

```

@@

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}

    do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000125931")

```

@@

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}

    do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000149294")

```

@@

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}

    do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000170577")

```

@@

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}

    do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000039068")

```

@@

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}

    do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000128567")

```

@@

- Combining label transfer and CNV inference we annotated 37/40 Wilms tumor samples in SCPCP000006.

- For the three remaining samples, we need to check why we do not have the output of `06_infercnv.R --HMM i3 --reference both`. 
It might be that we don't have enough normal cell for the reference.

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warnings = FALSE,
  knitr.digits.signif = FALSE
)
```

@@

### Output files

```{r}
# cell type annotation table
annotations_tsv <- file.path(result_dir, "SCPCP000006-annotations.tsv")
```

@@

```{r path_to_query}
sample_metadata_file <- file.path(repository_base, "data", "current", "SCPCP000006", "single_cell_metadata.tsv")

@@

# Create a data frames of all annotations
cell_type_df <- unique(metadata$scpca_sample_id) |>
  purrr::map(
    # For each sample_id, do the following:
    \(sample_id) {
      input_file <- file.path(
        result_dir,
        sample_id,
        glue::glue("06_infercnv_HMM-i3_{sample_id}_reference-both.rds")
      )

      # Read in the Seurat object
      if (file.exists(input_file)) {
        srat <- readRDS(input_file)

@@

          # cnv global estimation per chromosome
          has_cnv_chr1 = srat$has_cnv_chr1,
          has_cnv_chr2 = srat$has_cnv_chr2,
          has_cnv_chr3 = srat$has_cnv_chr3,
          has_cnv_chr4 = srat$has_cnv_chr4,
          has_cnv_chr5 = srat$has_cnv_chr5,
          has_cnv_chr6 = srat$has_cnv_chr6,
          has_cnv_chr7 = srat$has_cnv_chr7,
          has_cnv_chr8 = srat$has_cnv_chr8,
          has_cnv_chr9 = srat$has_cnv_chr9,
          has_cnv_chr10 = srat$has_cnv_chr10,
          has_cnv_chr11 = srat$has_cnv_chr11,
          has_cnv_chr12 = srat$has_cnv_chr12,
          has_cnv_chr13 = srat$has_cnv_chr13,
          has_cnv_chr14 = srat$has_cnv_chr14,
          has_cnv_chr15 = srat$has_cnv_chr15,
          has_cnv_chr16 = srat$has_cnv_chr16,
          has_cnv_chr17 = srat$has_cnv_chr17,
          has_cnv_chr18 = srat$has_cnv_chr18,
          has_cnv_chr19 = srat$has_cnv_chr19,
          has_cnv_chr20 = srat$has_cnv_chr20,
          has_cnv_chr21 = srat$has_cnv_chr21,
          has_cnv_chr22 = srat$has_cnv_chr22
        ) |>
          tibble::rownames_to_column("barcode") |>
          dplyr::mutate(sample_id = sample_id)
      }
    }
  ) |>
  # now combine all dataframes to make one big one
  dplyr::bind_rows()
```

@@

```{r fig.width=10, fig.height=4, out.width='100%'}
do_Feature_mean <- function(df, group.by, feature) {
  df <- df %>%
    group_by(sample_id, !!sym(group.by)) %>%
    summarise(m = mean(!!sym(feature)))

  p <- ggplot(df, aes(x = sample_id, y = !!sym(group.by), fill = m)) +
    geom_tile() +
    scale_fill_viridis_c() +
    theme_bw() +
    theme(text = element_text(size = 20)) +
    theme(axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 0.5)) +
    guides(fill = guide_colourbar(title = paste0(feature)))

@@

```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}
cell_type_df <- cell_type_df |>
  mutate(has_cnv_score = rowSums(cell_type_df[, grepl("has_cnv_chr", colnames(cell_type_df))])) |>
  mutate(has_cnv_score = case_when(
    has_cnv_score > params$cnv_threshold ~ "CNV",
    has_cnv_score <= params$cnv_threshold ~ "no CNV"
  ))

@@

```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}
# Define normal cells
# We first pick up the immune and endothelial cells annotated via the label transfer compartments under the condition that the predicted score is above the threshold
cell_type_df <- cell_type_df |>
  mutate(first.level_annotation = case_when(
    # assign normal/cancer based on condition
    compartment %in% c("fetal_nephron", "stroma") & has_cnv_score == "no CNV" ~ "normal",
    compartment %in% c("endothelium", "immune") & cell_type.score > params$predicted.celltype.threshold ~ "normal",
    compartment %in% c("fetal_nephron", "stroma") & has_cnv_score == "CNV" ~ "cancer",
    .default = "unknown"
  ))
```

@@

```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
ggplot(cell_type_df, aes(x = umap.umap_1, y = umap.umap_2, color = first.level_annotation), shape = 19, size = 1) +
  geom_point() +
  facet_wrap(facets = ~sample_id, ncol = 5) +
  theme_bw() +
  theme(text = element_text(size = 22))
```

@@

    # assign cancer cells based on condition
    cell_type_df$compartment %in% c("stroma") &
      cell_type_df$has_cnv_score == "CNV" ~ "cancer stroma",
    cell_type_df$compartment %in% c("fetal_nephron") &
      cell_type_df$has_cnv_score == "CNV" &
      cell_type_df$cell_type == "mesenchymal cell" ~ "blastema",
    cell_type_df$compartment %in% c("fetal_nephron") &
      cell_type_df$has_cnv_score == "CNV" &
      cell_type_df$cell_type != "mesenchymal cell" ~ "cancer epithelium",
    .default = "unknown"
  ))
```

```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
ggplot(cell_type_df[cell_type_df$first.level_annotation == "normal", ], aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation), shape = 19, size = 1) +
  geom_point() +
  facet_wrap(facets = ~sample_id, ncol = 5) +
  theme_bw() +
  theme(text = element_text(size = 22))
```

```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
ggplot(cell_type_df[cell_type_df$first.level_annotation == "cancer", ], aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation), shape = 19, size = 0.1) +
  geom_point() +
  facet_wrap(facets = ~sample_id, ncol = 5) +
  theme_bw() +
  theme(text = element_text(size = 22))
```

@@

```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
ggplot(cell_type_df, aes(x = umap.umap_1, y = umap.umap_2, color = second.level_annotation), shape = 19, size = 1) +
  geom_point() +
  facet_wrap(facets = ~sample_id, ncol = 5) +
  theme_bw() +
  theme(text = element_text(size = 22))
```

@@

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
for (i in 1:22) {
  print(do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = glue::glue("proportion_cnv_chr", i)))
}
```

@@

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000081237")
```

@@

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000110799")
```

@@

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000026025")
```

@@

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000163359")
```

@@

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000154096")
```

@@

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000125931")
```

@@

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000149294")
```

@@

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000170577")
```

@@

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000039068")
```

@@

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
do_Feature_mean(cell_type_df, group.by = "second.level_annotation", feature = "ENSG00000128567")
```

## Create annotation table for export

This section creates the cell type annotation table for export.

```{r}
annotations_table <- cell_type_df |>
  select(
    cell_barcode = barcode,
    scpca_sample_id = sample_id,
    tumor_cell_classification = first.level_annotation,
    cell_type_assignment = second.level_annotation
  ) |>
  mutate(
    # change cancer --> tumor, but keep the other labels
    tumor_cell_classification = ifelse(
      tumor_cell_classification == "cancer", "tumor", tumor_cell_classification
    ),
    cell_type_assignment = str_replace_all(
      cell_type_assignment,
      "cancer ",
      "tumor "
    )
  )

write_tsv(annotations_table, annotations_tsv)
```

Confirm how many samples we have annotations for:
```{r}
length(unique(annotations_table$scpca_sample_id))
```

@@

- Combining label transfer and CNV inference we annotated 38/40 Wilms tumor samples in SCPCP000006. For the two remaining samples, we need to investigate inferCNV further:
    - SCPCS000190 only has 1 normal cell, so we cannot specify a reference for inferCNV
    - SCPCS000203 does not run properly through inferCNV; cells are removed from the inferCNV inference, and why requires further investigation
